Bayesian Reconstruction of Coevolutionary Histories

### 2013-09-24 paper/article.tex
General progress

% Figures, tables, and captions
\usepackage{float}
\usepackage{booktabs}
\usepackage{caption}
\captionsetup{labelfont=bf,font=small,labelsep=period}
\usepackage{subcaption}
\usepackage[rightcaption]{sidecap}

@@

@@

\begin{titlepage}
\null
\vfil
\let\newpage\relax
\maketitle
\vfil
\centering
\begin{pspicture}(18,12)
\psset{unit=0.5cm,linewidth=0.2}
\psline[linecolor=blue](0,0)(10,10)
\psline[linecolor=blue](4,0)(2,2)
\psline[linecolor=blue](8,0)(4,4)
\psline[linecolor=blue](12,0)(14,2)
\psline[linecolor=blue](16,0)(8,8)
\psline[linecolor=red](1,0)(11,10)
\psline[linecolor=red,arrows=-o](17,0)(9,8)
\psline[linecolor=red,arrows=-o](13,0)(15,2)
\psline[linecolor=red](7,3)(14,3)
\psline[linecolor=red,arrows=<-](10,3)(14,3)
\psline[linecolor=red](10,0)(7,3)
\psline[linecolor=red,arrows=-o](9,0)(5,4)
\psline[linecolor=red,arrows=-o](4,1)(3,2)
\rput{135}(4,1){\LARGE\textcolor{red}{\textsf{\textbf{x}}}}
\psline[linecolor=red](18,0)(17,1)
\psline[linecolor=red,arrows=*-](16,1)(17,1)
\end{pspicture}
\vfil
\thispagestyle{empty}
\end{titlepage}

@@

Phylogenies, or evolutionary trees, can provide a valuable perspective to evolutionary processes. The topology of the tree depicts the set of speciation events that gave rise to those taxa and thus their ancestral relationships. Often, the lengths of branches indicate the number of substitutions (e.g., genetic mutations) accumulated or the amount of time passed at that particular lineage. By mapping traits such as physical attributes or geographic locations onto a phylogeny, we can use the tree to learn about the mechanisms under which these traits change.

@@

Often we are interested in only some of the parameters (e.g., just the tree topology), in which case we can calculate the posterior probability of specific values for those parameters by integrating over all possible values of the remaining \enquote{nuisance} parameters (branch lengths, mutation rate, etc.). Due to the complexity of the models, the resulting multi-dimensional integral cannot be evaluated directly but instead approximated via a \ac{MCMC} .

The Metropolis--Hastings algorithm \parencites{Metropolis:1953}{Hastings:1970} is the most common implementation of \ac{MCMC} utilized in phylogenetic inference

@@

A primary objective of parsimony approaches to the cophylogeny problem is developing heuristics to find optimal reconstructions quickly. For example,\textcite{Charleston:1998} proposed the \enquote{jungle} concept and the program Jane utilizes a genetic algorithm to find optimal timings \parencite{Conow:2010}. This problem becomes much less of a concern in an \ac{MCMC} framework, because the parameter space is being explored as opposed to maximized (although it is still critical to employ effective operators for efficient mixing of the chain). However, it is important to note that \ac{MCMC} carries a much greater computational expense.

\begin{figure}
\centering
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0.5)(0.5,0.5)
\psline[linecolor=blue](0.5,0.3)(1,0.3)
\psline[linecolor=blue](0.5,0.7)(1,0.7)
\psline[linecolor=blue](0.5,0.3)(0.5,0.7)
\psline[linecolor=red](0.4,0.4)(1,0.4)
\psline[linecolor=red](0.4,0.8)(1,0.8)
\psline[linecolor=red](0.4,0.4)(0.4,0.8)
\psline[linecolor=red,arrows=-o](0,0.6)(0.4,0.6)
\end{pspicture}
\caption{cospeciation}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=red](0,0.1)(1,0.1)
\psline[linecolor=red,arrows=*-](0.5,0.1)(0.5,0.3)
\psline[linecolor=red](0.5,0.3)(1,0.3)
\end{pspicture}
\caption{duplication}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=blue](0,0.4)(1,0.4)
\psline[linecolor=red](0,0.1)(1,0.1)
\psline[linecolor=red](0.5,0.1)(0.5,0.5)
\psline[linecolor=red,arrows=->,arrowsize=0.1](0.5,0.1)(0.5,0.35)
\psline[linecolor=red](0.5,0.5)(1,0.5)
\end{pspicture}
\caption{host-switch}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=red](0,0.1)(0.8,0.1)
\rput(0.8,0.1){\large\textcolor{red}{\textsf{x}}}
\end{pspicture}
\caption{loss}
\end{subfigure}
\caption{Diagrams of the four cophylogenetic events considered in my model, where the host organism's phylogeny is in blue and the symbiont phylogeny in red. All events are branch events (i.e., they can occur at any point along a branch) except cospeciation, which is a nodal event (and thus can occur only when the host speciates).}
\end{figure}

@@

\paragraph*{Casey study: cycad--weevil}

>>>>

% Figures, tables, and captions
\usepackage{float}
\usepackage{booktabs}
\usepackage[hypcap]{caption}
\captionsetup{labelfont=bf,font=small,labelsep=period}
\usepackage{subcaption}
\usepackage[rightcaption]{sidecap}
\usepackage{fancyref}

@@

% Reusable figures
\newcommand{\pscophylogeny}{
\begin{pspicture}(18,12)
\psset{unit=0.5cm,linewidth=0.2}
\psline[linecolor=blue](0,0)(10,10)
\psline[linecolor=blue](4,0)(2,2)
\psline[linecolor=blue](8,0)(4,4)
\psline[linecolor=blue](12,0)(14,2)
\psline[linecolor=blue](16,0)(8,8)
\psline[linecolor=red](1,0)(11,10)
\psline[linecolor=red,arrows=-o](17,0)(9,8)
\psline[linecolor=red,arrows=-o](13,0)(15,2)
\psline[linecolor=red](7,3)(14,3)
\psline[linecolor=red,arrows=<-](10,3)(14,3)
\psline[linecolor=red](10,0)(7,3)
\psline[linecolor=red,arrows=-o](9,0)(5,4)
\psline[linecolor=red,arrows=-o](4,1)(3,2)
\rput{135}(4,1){\LARGE\textcolor{red}{\textsf{\textbf{x}}}}
\psline[linecolor=red](18,0)(17,1)
\psline[linecolor=red,arrows=*-](16,1)(17,1)
\end{pspicture}
}

@@

\begin{titlepage}
\null
\vfil
\let\newpage\relax
\maketitle
\vfil
\centering
\pscophylogeny
\vfil
\thispagestyle{empty}
\end{titlepage}

@@

Phylogenies, or evolutionary trees, can provide a valuable perspective to evolutionary processes. The topology of the tree depicts the set of speciation events that gave rise to those taxa and thus their ancestral relationships. Often, the lengths of branches indicate the number of substitutions (e.g., genetic mutations) accumulated or the amount of time passed at that particular lineage. By mapping traits such as physical attributes or geographic locations onto a known phylogeny, we can use the tree to learn about the mechanisms under which these traits change.

@@

Often we are interested in only some of the parameters (e.g., just the tree topology), in which case we can calculate the posterior probability of specific values for those parameters by integrating over all possible values of the remaining \enquote{nuisance} parameters (branch lengths, mutation rate, etc.). Due to the complexity of the models, the resulting multi-dimensional integral cannot be evaluated directly but instead approximated via \ac{MCMC}.

The Metropolis--Hastings algorithm \parencites{Metropolis:1953}{Hastings:1970} is a common implementation of the \ac{MCMC} method in phylogenetic inference

@@

A primary objective of parsimony approaches to the cophylogeny problem is developing heuristics to find optimal reconstructions quickly. For example, \textcite{Charleston:1998} proposed the \enquote{jungle} concept and the program Jane utilizes a genetic algorithm to find optimal timings \parencite{Conow:2010}. This problem becomes much less of a concern in an \ac{MCMC} framework, because the parameter space is being explored as opposed to maximized (although it is still critical to employ effective operators for efficient mixing of the chain). However, it is important to note that \ac{MCMC} carries a much greater computational expense.

\begin{figure}
\centering
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0.5)(0.5,0.5)
\psline[linecolor=blue](0.5,0.3)(1,0.3)
\psline[linecolor=blue](0.5,0.7)(1,0.7)
\psline[linecolor=blue](0.5,0.3)(0.5,0.7)
\psline[linecolor=red](0.4,0.4)(1,0.4)
\psline[linecolor=red](0.4,0.8)(1,0.8)
\psline[linecolor=red](0.4,0.4)(0.4,0.8)
\psline[linecolor=red,arrows=-o](0,0.6)(0.4,0.6)
\end{pspicture}
\caption{cospeciation}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=red](0,0.1)(1,0.1)
\psline[linecolor=red,arrows=*-](0.5,0.1)(0.5,0.3)
\psline[linecolor=red](0.5,0.3)(1,0.3)
\end{pspicture}
\caption{duplication}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=blue](0,0.4)(1,0.4)
\psline[linecolor=red](0,0.1)(1,0.1)
\psline[linecolor=red](0.5,0.1)(0.5,0.5)
\psline[linecolor=red,arrows=->,arrowsize=0.1](0.5,0.1)(0.5,0.35)
\psline[linecolor=red](0.5,0.5)(1,0.5)
\end{pspicture}
\caption{host-switch}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\centering
\begin{pspicture}(1,1)
\psline[linecolor=blue](0,0)(1,0)
\psline[linecolor=red](0,0.1)(0.8,0.1)
\rput(0.8,0.1){\large\textcolor{red}{\textsf{x}}}
\end{pspicture}
\caption{loss}
\end{subfigure}
\caption{Diagrams of the four cophylogenetic events considered in my model, where the host organism's phylogeny is in blue and the symbiont phylogeny in red. All events are branch events (i.e., they can occur at any point along a branch) except cospeciation, which is a nodal event (and thus can occur only when the host speciates).}
\label{fig:events}
\end{figure}

\begin{figure}
\centering
\pscophylogeny
\caption{An example reconstruction of a coevolutionary history, where notation follows \Fref{fig:events}.}
\label{fig:cophylogeny}
\end{figure}

@@

\paragraph*{Case study: cycad--weevil}
